---
title: "Coyotepark"
output: html_document
---

# Data Loading

```{r loading_data}
library(here)
library(tidyverse)
library(ggplot2)
library(ggmap)
library(mapdata)
library(maps)
library(mapproj)
library(glmmTMB)
library(ggeffects)
library(lme4)
library(DHARMa)
library(performance)
library(sf)
library(spatstat)
library(tigris)
library(rnaturalearth)
library(patchwork)

hookworm <- read.csv(here("Data","hw_data_for_gabby.csv"))
animal_info <- read.csv(here("Data","carcasses_5-7-25.csv"))
#testing gabby github issue
```

First we have to combine our hookworm dataset and the animal info.

```{r combining_datasets}
#we know we'll want one datasheet that has both the hookworm data for each coyote, but also the metadata (ex: lat/long, animal size, etc). We can combine our datasets by either the 'Sample.ID' column or the 'Prep.Number' column, but in order to work, the entries have to match EXACTLY, and I noticed in the hookworm dataframe, the Sample.ID entries are lowercase, whereas in the animal_info dataframe, they're uppercase. So I'm going to combine the two sheets using the Prep.Number column. 

coyote_combo<-merge(hookworm, animal_info, by.x = "Prep.Number", by.y = "Prep_Number", all.x = TRUE)
View(coyote_combo)

#yay, that worked! now i'm going to tidy the datasheet and remove columns that we don't need (not necessary, but makes it simpler to work with imo)

coy_trim <- coyote_combo %>%
  select(c(Prep.Number, Sample.ID, Hookworm..F., Hookworm..M., Total.Hookworms, Date_of_death, County_approximate._ocation, City_approximate_location, Latitude, Longitude, Sex, Age, Total_length_.mm., Weight_.g., Body_condition_.1.5.)) %>% #this long list is all the variables I want to keep from the dataset!
    janitor::clean_names() #this makes all the labelling tidy + consistent 
View(coy_trim)

#unfortunately, there are two coyotes where we don't have lat/lon data from, so i'm going to remove them since we we won't be able to use them in our analysis
coy_trim <- coy_trim %>% 
  filter(!is.na(latitude))
View(coy_trim)

#cool! now we have a dataframe with 94 entries that we can use to run our analysis.
```

Before we do any specific analyses, let's take a basic look at some trends.

```{r data_vis}
#just for fun, i want to look at some potential trends!
#need to make sure R views lat/long as continuous data
coy_trim$latitude <- as.numeric(coy_trim$latitude)
coy_trim$longitude <- as.numeric(coy_trim$longitude)

#some basic plots
ggplot(coy_trim, aes(x=latitude, y=total_hookworms)) + #denotes x and y
  geom_point() + #denotes scatterplot
  geom_smooth(formula = y ~ x, method = "lm") #adds simple line of best fit

ggplot(coy_trim, aes(x=longitude, y=total_hookworms)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm")

#doesn't look like there are strong lat/lon trends, which is fine!

#now i want to plot all our points on a map
#base map layer
bounds<-c(left=-122.7, bottom=47.1, right=-122, top=48.5)
mid_x <- (bounds["left"] + bounds["right"]) / 2
bottom_y <- bounds["bottom"]
top_y <- bounds["top"]
mid_y <- (bottom_y + top_y) / 2
base_map <- get_stadiamap(bounds, zoom = 12, maptype = "stamen_terrain_background", extent="device")
ggmap(base_map)
#plot map
coyote_map <- ggmap(base_map) +
  geom_point(data=coy_trim, aes(x=longitude, y=latitude), 
             position=position_jitter(width=0.001, height=0.001), 
             shape=21, size=3, stroke=0.5)
coyote_map

#cool! we can see from this map that our coyotes are spread around the puget sound area, not just in seattle
```

#Analysis

Where can we take this analysis?
1. Making a continuous gradient map of risk across the whole system (like Chelsea's schisto paper)
2. Analyzing disease incidence along an urban-rural gradient
3. Mapping each coyote to a nearby park within the city of Seattle

## Yasmine Population Density Data

```{r uploading_yas_popdata}
#first loading in the data
popdens <- read.csv(here("Data","carc_pts_with_covars.csv"))

#next trimming it to ONLY include the population data and the prep #, and remove all non-coyotes
popdens_2 <- popdens %>%
  filter(species == "Coyote") %>%
  select(c(prep_number, popdens_1000m, popdens_2000m, popdens_3000m))

#now merging these with our existing sheet (yay)
coy_final<-merge(coy_trim, popdens_2, by.x = "prep_number", by.y = "prep_number", all.x = TRUE)

#making sure R reads columns correctly
coy_final$latitude <- as.numeric(coy_final$latitude)
coy_final$longitude <- as.numeric(coy_final$longitude)
coy_final$weight_g <- as.numeric(coy_final$weight_g)

#dropping coyotes with N/A weight or age
coy_final <- coy_final %>% 
  filter(!is.na(weight_g)) %>%
  filter(!is.na(age))
```

```{r yas_models}
ggplot(coy_final, aes(x=popdens_3000m, y=total_hookworms)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm")

#looks to be a slightly negative trend here! but now we need to model it...
#to decide what kind of model to use, i'll make a histogram to see the shape of the data

hist(coy_final$total_hookworms)

#see how this histogram has a TON of values on the low end, and very few on the high end? this is a classic shape for parasite data (many hosts have no/low infections, a few hosts have severe infections) which you might remember from parasite ecology, called an aggregated distribution. it's much better to use a distribution called a "negative binomial" distribution, which takes into account the way count/parasite data is usually shaped.

# we have to pick out our predictors! crash course incoming BUT... think of predictors as the variables that will be important in our model. here we want to test what influences hookworm burden. predictors of interest will be: lat, long, age and size of coyote, and popdens. 

#more complicated models have trouble "converging", so that's something i'll work through below!

#simple model, no weight, converges :-)
yas1 <- glmmTMB(total_hookworms ~ popdens_3000m + (1|latitude) + (1|age), data=coy_final, family=nbinom12)
summary(yas1)

#model including weight doesn't converge :-(
yas2 <- glmmTMB(total_hookworms ~ popdens_3000m + (1|latitude) + (1|age) + (1|weight_g), data=coy_final, family=nbinom12)
summary(yas2)

#what if we try length?
yas3 <- glmmTMB(total_hookworms ~ popdens_3000m + (1|latitude) + (1|age) + (1|total_length_mm), data=coy_final, family=nbinom12)
summary(yas3)

#yay, that works much better! i think weight is tricky for this model because it has HUGE variation, whereas the total length has less. both measure the animal size, so i'm not sure one is better or worse- talk to yasmine?
```

```{r model_diagnostics}
models <- list(yas1, yas3)
model_names <- c("yas1", "yas3")

#model residuals and diagnostics
#these are measures of how well the model fits the data, and if there are any significant outliers! this code will pop out a bunch of graphs. we hope that NONE of them are significant, because any significance here means our models are not good
residual_summaries_richness <- map2_df(models, model_names, function(model, name) {
  sim <- simulateResiduals(fittedModel = model, plot = FALSE, re.form=NULL)
  data.frame(
    model = name,
    dispersion = testDispersion(sim)$p.value,
    uniformity = testUniformity(sim)$p.value,
    outliers = testOutliers(sim)$p.value,
  )
})
#yay, no significance!

#AIC
#aic measures the "loss" of information that happens when you create a model. for this, a lower number is better. 
AIC(yas1)
AIC(yas3)
#yas3 has the lower AIC! so we'll stick with that one moving forward
```

```{r yas_plots}
#this first line tells your plot which model to include!
yas3_mplot <- ggpredict(yas3,c("popdens_3000m"))

#change plot
yas3_plot <- plot(yas3_mplot, show_data = TRUE) + 
  xlab("human population density") +
  ylab("predicted hookworm count")

#display plot
yas3_plot

#new plot
sofplot <-plot(yas3_mplot, show_data = TRUE, color = "#b73779") +
  xlab("human population density") +
  ylab("predicted hookworm count") +
  theme_classic() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Main Title", subtitle = "Optional Subtitle", caption = "Data source info") 
sofplot
#scale_fill_manual(values = "#b73779") ?
#geom_line(color = "#b73779") + ?

#cool!!! we are seeing clearly in this graph that the risk of coyote hookworm is HIGHER when there are LESS human population density. we can talk about how to interpret this soon!!!
```

## Mapping Coyotes to Parks
```{r seattle}
# first we should trim our data to include only the coyotes within the seattle metro area!!
#northerly bound: 145th st, or 47.734110
#southerly bound: kind of haphazard, but i did my best estimate at 47.495584

coy_seattle <- coy_final %>%
  filter(latitude < 47.734110) %>%
  filter(latitude > 47.495584)

#ok... that only leaves us with 12 data points, not enough for an analysis! so im gonna expand our search frame up to about everett and down to seatac.

coy_metro <- coy_final %>%
  filter(latitude < 48.017612) %>%
  filter(latitude > 47.434439)

#37 data points... still not a lot but more to work with!

#mapping!
bounds<-c(left=-122.7, bottom=47.4, right=-122, top=48.02)
mid_x <- (bounds["left"] + bounds["right"]) / 2
bottom_y <- bounds["bottom"]
top_y <- bounds["top"]
mid_y <- (bottom_y + top_y) / 2
base_map <- get_stadiamap(bounds, zoom = 12, maptype = "stamen_terrain_background", extent="device")
ggmap(base_map)
#plot map
coyote_sea_map <- ggmap(base_map) +
  geom_point(data=coy_metro, aes(x=longitude, y=latitude), 
             position=position_jitter(width=0.001, height=0.001), 
             shape=21, size=3, stroke=0.5)
coyote_sea_map

#next steps: see those points on a map, connect them to closest parks, see if park risk factor is a useful factor in a model!

```

```{r seattle_parks}
View(coy_metro)

#i can't think of an automatic way to do this, so im gonna manually assign each coyote to a park by creating a vector in the same order

parks <- c("ballinger", "brightwater", "north creek", "hans jensen", "paine field", "bridle trails", "paine field", "picnic point", "dakota", "paine field", "dakota", "hans jensen", "north creek", "hans jensen", "paine field", "hans jensen", "picnic point", "paine field", "paine field", "paine field", "squak mountain", "paine field", "discovery", "paine field", "paine field", "paine field", "blyth", "paine field", "newport bluffs", "magnuson", "paine field", "paine field", "paine field", "picnic point", "mercer slough", "paine field", "discovery")

#next add this column to datasheet
coy_metro$parks <- parks

#lets do some data visualization!
ggplot(data= coy_metro, aes(x=parks, y=total_hookworms)) +
  geom_boxplot()

#promising... now lets try a model
yas4 <- glmmTMB(total_hookworms ~ parks, data=coy_metro, family=nbinom12)
summary(yas4)
#even the simplest of these models doesn't converge, maybe because there are some parks that only have 1 replication. let's drop those

coy_metro_fil <- coy_metro %>%
  filter(parks != "ballinger") %>%
  filter(parks != "brightwater") %>%
  filter(parks != "bridle trails") %>%
  filter(parks != "squak mountain") %>%
  filter(parks != "blyth") %>%
  filter(parks != "newport bluffs") %>%
  filter(parks !="magnuson") %>%
  filter(parks != "mercer slough")

coy_metro_fil$parks <- as.character(coy_metro_fil$parks)

#data vis
ggplot(data= coy_metro_fil, aes(x=parks, y=total_hookworms)) +
  geom_boxplot()

#making an anova to compare multiple groups
coy_anova <- aov(total_hookworms ~ parks, data = coy_metro_fil)
hist(coy_anova$residuals)
qqPlot(coy_anova$residuals, id = FALSE)
summary(coy_anova)
#not significant, and there are so few points that its difficult to make big assumptions about this data!! :-(

#tukey compares each group individually
tukey <- TukeyHSD(coy_anova)
tukey
#here we see no singificant difference between any of the parks!!
```

## Gradient Maps

```{r basic_map}
#ok this is gonna be the hardest part i fear!! must upload the map using a different program. for reference, i am following mainly these resources:
# https://www.andrewheiss.com/blog/2023/07/28/gradient-map-fills-r-sf/
# https://bartholomatt.net/posts/census_maps_part2/
# https://library.virginia.edu/data/articles/spatial-r-using-sf-package

#first loading our base map!
# Geographic data from the US Census
options(tigris_use_cache = TRUE)
Sys.setenv(TIGRIS_CACHE_DIR = "maps")
wa_crs <- st_crs("EPSG:4326")

#state
wa_state <- states(cb = TRUE, resolution = "500k", year = 2022) %>% 
  filter(STUSPS == "WA") %>% 
  st_transform(wa_crs)
#counties
wa_counties <- counties(state = "53", cb = TRUE, resolution = "500k", year = 2022) %>%
  st_transform(wa_crs)
#skipping adding water bc it was really confusing to me
wa <-  'WA'
wac <- c("King", "Pierce", "Snohomish", "Island", "Skagit", "Whatcom", "San Juan", "Mason", "Jefferson", "Clallam", "Kitsap", "Thurston")
wa_water <- area_water(wa, wac)
landmass_data = counties(cb = TRUE, state = wa)

#transform our coyote dataframe into sf data
coy_sf <- st_as_sf(as.data.frame(coy_final), coords = c("longitude", "latitude"), remove = FALSE, crs = st_crs(wa_state))
View(coy_sf)

st_crs(wa_state) == st_crs(coy_sf)

#plot
plot_initial <- ggplot() +
  geom_sf(data = wa_state, fill = "grey20") +
  geom_sf(data = coy_sf, size = 0.5, color = "grey50") +
  #geom_point(data = coy_final, aes(x=longitude, y=latitude), size = 0.8, color = "grey50") +
  coord_sf(crs = wa_crs)
plot_initial

#cool! can i now trim this map to a smaller area of interest for us?
plot_initial <- ggplot() +
  geom_sf(data = wa_state, fill = "grey20") +
  geom_sf(data = coy_sf, size = 0.5, color = "lightgrey") +
  xlim(123, 121.6) +
  ylim(47, 49) +
  theme_void()
plot_initial
#yippeee

#here is the main issue: the coyote coordinates only plot properly when in EPSG:4326. but because this is in degrees not m, i cant do the below grid transformations. how do i make the coyote plot accurately in m?

```

```{r grid_gradient}
#need to reset state/coy data in projection system that uses m
coy_sf2 <- st_set_crs(coy_sf, "EPSG:32149")
coy_sf2 <-  st_transform(coy_sf, "EPSG:32149")
View(coy_sf2)

wa_state2 <- st_set_crs(wa_state, "EPSG:32149")
wa_state2 <-  st_transform(wa_state, "EPSG:32149")

plot_test <- ggplot() +
  geom_sf(data = wa_state2, fill = "grey20") +
  geom_sf(data = coy_sf2, size = 0.5, color = "lightgrey") +
  xlim(320000, 400000) +
  ylim(190000, 400000)
plot_test

#i literally cant believe it finally worked lol

#ok now the weird stuff
# Spit the state area into a 20x20 grid
wa_grid <- wa_state2 %>% 
  st_make_grid(n = c(50, 50))

wa_grid_map <- st_intersection(wa_state2, wa_grid) %>% 
  st_as_sf() %>% 
  mutate(grid_id = 1:n())

#look at plot
ggplot() +
  geom_sf(data = wa_grid_map) +
  #geom_sf(data = wa_grid, alpha = 0.3) +
  geom_sf(data = wa_water, fill = 'white',color = NA, aes(geometry = geometry)) +
  #theme(panel.background = element_rect(fill = '#cce6ff', color = NA)) +
  xlim(320000, 400000) +
  ylim(190000, 400000) 
  #theme_void()

#can we make a grid gradient representing number of hookworms/coyotes?
coyotes_per_grid_box <- wa_grid_map %>% 
  st_join(coy_sf2) %>% 
  group_by(grid_id) %>% 
  summarize(total = sum(!is.na(total_hookworms)))

#this is how we plot it
hookworm_grid <- ggplot() +
  geom_sf(data = wa_state2, fill = NA, color = "black", linewidth = 0.25) +
  geom_sf(data = coyotes_per_grid_box, aes(fill = total), color = "black") +
  geom_sf(data = wa_water, fill = '#cce6ff',color = NA, aes(geometry = geometry)) +
  geom_sf(data = coy_sf2, size = .3, color="grey80") +
  scale_fill_viridis_c(option = "magma", guide = "none") +
  xlim(320000, 400000) +
  ylim(190000, 400000) +
  theme_void()

hookworm_grid

#OMG IT WORKED FINALLY :-P probably need to add a legend to this!

#what about gradient??
# Convert the campground coordinates to a ppp object with a built-in window
coy.ppp <- as.ppp(coy_sf2$geometry, W = as.owin(wa_state2)) 

# Check to see if it worked
plot(coy.ppp)

# Create a stars object of the density of coyotes
density_coy_stars <- stars::st_as_stars(density(coy.ppp, dimyx = 300))

# Check to see what it looks like
plot(density_coy_stars)

wa_coy_density <- st_as_sf(density_coy_stars) %>%
  st_set_crs("EPSG:32149")

plot_density <- ggplot() +
  geom_sf(data = wa_coy_density, aes(fill = v), color = NA) +
  geom_sf(data = wa_state2, fill = NA, color = "black", linewidth = 0.25) +
  scale_fill_viridis_c(option = "magma", guide = "none") +
  geom_sf(data = wa_water, fill = '#cce6ff',color = NA, aes(geometry = geometry)) +
  geom_sf(data = coy_sf2, size = 0.5, color = "lightgrey") +
  xlim(320000, 400000) +
  ylim(190000, 400000)
plot_density
```
